\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-03-18 docs/parco13/abstract.tex
Remove two undefined references
Minor changes

\usepackage{amsmath,amssymb}
\usepackage{cite}
\usepackage{color}
\usepackage{graphicx}
\usepackage{url}
\usepackage{xspace}

@@

@@

Consequently, it is not clear a priori how to accelerate PDE solvers
that use adaptive mesh refinement.
%, especially when working with
%unstructured meshes \comment{This sounds we are now going to talk
%  about unstructured meshes.}.
% CB: Good point, took it out.
Of course, it was realized early that it helps to aggregate degrees of
freedom (DOF) at the element level, as has been done with high-order
spectral element \cite{TufoFischer99}, low order continuous Galerkin via methods that accumulate many elements simultaneously \cite{knepleyterrel:2013},
%finite volume (element?)
or discontinuous Galerkin \cite{HesthavenWarburton02}
methods.  GPU implementations of the latter have been proposed recently
\cite{KlocknerWarburtonBridgeEtAl09, BursteddeGhattasGurnisEtAl10}.
The finite volume method has typically been implemented using a single degree
of freedom per cell on
% either
structured
\cite{ppm, clawpack}
or unstructured meshes
\cite{ddfv, ausm+, openfoam},
where higher order methods have been constructed by widening the stencil, for instance in WENO methods
\cite{Shu:2009bi}.
% \comment{Are you looking for
%   references to parallel AMR codes?  Or more general AMR codes based
%   on the FVM?}  \todo{No, I was looking for unstructured FV
%   references.  Would you have any, and can we add the block-AMR
%   references later?} \comment{But most finite volume methods
%   (structured or unstructured) only have a single degree of freedom
%   per mesh cell (``element''?)  Only recently have people increased
%   the order by extending the stencil (without increase the DOFs per
%   mesh cell).}
% CB: Can you fix some of the above citations?

@@

\section{Numerical results}

\begin{figure}
\begin{center}
\includegraphics[width=.4\columnwidth]{quadtree}
\hspace{.05\columnwidth}
\includegraphics[width=.4\columnwidth]{sphere_plot167}
\end{center}
\caption{Numerical results for solving the advection equation.  Left:  Unit
square (single quadtree) with a twisting velocity field.  Right: Spherical ball
with a rotational velocity field constructed from two mapped quadtrees.  In
both cases the concentration is color-coded with a sharp gradient shown in red.
The adaptive mesh refinement follows the location of the gradient (the patches
are not shown where they become too fine for display).  Here we use $m = 8$
cells per \clawpack patch.}
\label{results2d}
\end{figure}

The results in this section are preliminary and will be extended in the final
version.  We have so far exercised \forestclaw for two 2D manifolds, namely the
unit square and a sphere composed of two mapped squares.  On both we solve the
advection equation with a given velocity field.  We advect a concentration
field with a sharp gradient which we track dynamically with adaptive mesh
refinement.  We observe that the refinement follows the front and allows for an
accurate representation of its location over time; see Figure~\ref{results2d}.

@@

The authors acknowledge valuable discussion with Randy LeVeque, Marsha Berger,
and Hans-Petter Langtangen.  We also acknowledge David Ketcheson and the KAUST
sponsored HPC$^3$ numerics workshop at which the initial phases of this project
were first discussed.
The leaf/patch paradigm was independently presented
by B.\ as part of a talk at the SCI Institute, Utah, in July 2011.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{cite}
\usepackage{color}
\usepackage{graphicx}
\usepackage{url}
\usepackage{xspace}
\usepackage{graphicx}

@@

% \newcommand{\plotbox}[1]{\fbox{#1}}
\newcommand{\plotbox}[1]{#1}

\newcommand{\Fig}[1]{Figure \ref{fig:#1}}

@@

Consequently, it is not clear a priori how to accelerate PDE solvers
that use adaptive mesh refinement.
%, especially when working with
%unstructured meshes \comment{This sounds we are now going to talk
%  about unstructured meshes.}.
% CB: Good point, took it out.
Of course, it was realized early that it helps to aggregate degrees of
freedom (DOF) at the element level, as has been done with high-order
spectral element \cite{TufoFischer99}, low order continuous Galerkin via methods that accumulate many elements simultaneously \cite{knepleyterrel:2013},
%finite volume (element?)
or discontinuous Galerkin \cite{HesthavenWarburton02}
methods.  GPU implementations of the latter have been proposed recently
\cite{KlocknerWarburtonBridgeEtAl09, BursteddeGhattasGurnisEtAl10}.
The finite volume method has typically been implemented using a single degree
of freedom per cell on
% either
structured
\cite{ppm, clawpack}
or unstructured meshes
\cite{li:2010, he-hu:2008, do-om:2005, li-sh-sv-va:2007, openfoam},
where higher order methods have been constructed by widening the stencil, for instance in WENO methods
\cite{Shu:2009bi}.
% \comment{Are you looking for
%   references to parallel AMR codes?  Or more general AMR codes based
%   on the FVM?}  \todo{No, I was looking for unstructured FV
%   references.  Would you have any, and can we add the block-AMR
%   references later?} \comment{But most finite volume methods
%   (structured or unstructured) only have a single degree of freedom
%   per mesh cell (``element''?)  Only recently have people increased
%   the order by extending the stencil (without increase the DOFs per
%   mesh cell).}
% CB: Can you fix some of the above citations?

@@

\section{Numerical results}
Below, we provide two examples that demonstrate the \forestclaw code.
The numerical results to date have been designed to verify that
interface layer between \pforest and \forestclaw is sufficiently
flexible and robust.  Of particular importance was ensuring that all
ghost cell ghost cell transfers (including averaging and
interpolation) have been implemented correctly.
The basic \clawpack algorithm and corresponding code have been
thoroughly tested, and so needed no further verification here.

In both sets of numerical results, we solve a scalar advection
equation, given by
\begin{equation*}
q_t + ({\bf u}\cdot \nabla) q = 0
\end{equation*}
\noindent
where the velocity field ${\bf u} = (u(\xi,\eta,t), v(\xi,\eta,t)$ is
a prescribed function of computational coordinates.  The relevant
numerical parameters that are set in each case include the size of the
patch on each leaf ($m=8$ in each case), and the minimum and maximum
levels, which in turn fix minimum and maximum effective grid
resolutions.

In both examples, a patch is tagged for refinement if the
difference between its maximum and minimum values exceed a prescribed
threshold.  A family of patches (4 sibling patches in 2d) is coarsened
if a coarsened parent patch did not meet the criteria for refinement.

\noindent
{\bf Example 1 : } In this example, an initial concentration field $q$
is set to 0 in the left half of a computational square and to 1 in the
right half.  A time dependent flow field is prescribed that distorts
the interface between the 0 and 1 concentration values.  The minimum
and maximum levels of refinement were set to 3 and 6, respectively.
This results in a minimum resolution of $64 \times 64$ and a maximum
resolution of $512 \times 512$. \Fig{results2d} shows the results
at an intermediate time step.

% \begin{figure}
% \begin{center}
% \plotbox{\includegraphics[height=0.485\textwidth,clip=true,trim=3cm 0cm 2cm 0cm]{quadtree.png}} \hfil
% \plotbox{\includegraphics[height=0.485\textwidth,clip=true,trim=6cm 0cm 7cm 0cm]{sphere_plot167.png}}
% \end{center}
% \caption{Results for scalar advection using \forestclaw.}
% \label{fig:results}
% \end{figure}

\noindent {\bf Example 2 : } This example, we demonstrate the
multiblock functionality of \forestclaw by considering flow on a
sphere.  The sphere mapping we use consists of two quadtrees, each
defined in the computational space $[-1,1] \times [-1,1]$
\cite{ca-he-le:2008, be-ca-he-le:2009}.  Each quadtree is mapped to cover one hemisphere
of the sphere.  The initial conditions were set to 0 in one half of
the sphere, and 1 in the other half, where the halves are not
necessarily aligned with the equator of the mapping.  The flow field
${u}$ simulates rigid body rotation.  We show the results at an intermediate time in
\Fig{results2d}.

\begin{figure}
\begin{center}
\plotbox{\includegraphics[width=.4\columnwidth]{quadtree}} \hfil
% \hspace{.05\columnwidth}
\plotbox{\includegraphics[width=.4\columnwidth]{sphere_plot167}}
\end{center}
\caption{Numerical results for solving the advection equation.  Left:  Unit
square (single quadtree) with a twisting velocity field.  Right: Spherical ball
with a rotational velocity field constructed from two mapped quadtrees.  In
both cases the concentration is color-coded with a sharp gradient shown in red.
The adaptive mesh refinement follows the location of the gradient (the patches
are not shown where they become too fine for display).  Here we use $m = 8$
cells per \clawpack patch.}
\label{fig:results2d}
\end{figure}

The results in this section are preliminary and will be extended in the final
version.  We have so far exercised \forestclaw for two 2D manifolds, namely the
unit square and a sphere composed of two mapped squares.  On both we solve the
advection equation with a given velocity field.  We advect a concentration
field with a sharp gradient which we track dynamically with adaptive mesh
refinement.  We observe that the refinement follows the front and allows for an
accurate representation of its location over time; see \Fig{results2d}.

@@

The authors acknowledge valuable discussion with Randy LeVeque, Marsha
Berger, and Hans-Petter Langtangen.  We also acknowledge David
Ketcheson and the KAUST sponsored HPC$^3$ numerics workshop at which
the initial phases of this project were first discussed.  The second
author would like to also acknowledge the Isaac Newton Institute
(Cambridge, UK), where much of the preliminary development work for
\forestclaw was done. The leaf/patch paradigm was independently
presented by B.\ as part of a talk at the SCI Institute, Utah, in July
2011.
